---
title: "DURIN 2026 Thermal Tolerance Calluna Model 3"
author: "Sonya R Geange and Gaëlle Poulat"
format: html
editor: visual
---

# Import data

```{r}
#| label: setup
#| message: false 

# installing / loading packages
library(readxl) # Enables excel file import
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(janitor) # Has tools for data cleaning 
library(lubridate) # Date and time formatting
library(readr) # Data importation functions
library(stringr) # String and character text
library(future) # Using severals cores at the same time
library(ggcorrplot)
library(emmeans)
library(car) # for the Anova
library(dplyr)
library(future)
```

```{r}
data <- read.csv2("C:/Users/poula/Documents/M2_PAMEQ/Stage M2/Analyse statistique/thermotolerance/data/clean_data/Seasonal Calluna Thermal Tolerance and Plant Functional Traits.csv")
```

```{r}
#| label: Create-united-columns-traits-data-frame

traits_data <- data |> 
  filter(proj_2025 == "Traits") |> 
  filter(!(month == "March" &
             age_class %in% c("Mature", "Building"))) |>  # We did not keep MAT and BUI in March because we didn't estimate T50.
  dplyr::select(site_name, species, month, age_class, plant_nr, leaf_age, height_flower, height_vegetative, leaf_area, wet_mass, dry_mass, SLA, LDMC)
  
united_columns_traits_plante <- unite(
  traits_data,
  col = "new_id",
  site_name, species, month,  age_class, plant_nr,leaf_age,
  sep = "_",
  remove = TRUE
) |>
  na.omit()
```

## Effect of leaf age on traits

```{r}
#| label: Create-data-frame-May-young-old

May_young_old <- traits_data |> 
  filter(month=="May")|> 
  drop_na() # some leaves were not harvested for some plants

```

```{r}
#| label: PCA


library(FactoMineR)
library(factoextra)

traits_old_young <- May_young_old[, c(
  "SLA",
  "leaf_area",
  "LDMC",
  "height_vegetative"
)]

res.pca <- PCA(
  traits_old_young,
  scale.unit = TRUE,
  graph = FALSE
)

# Graphique des individus
fviz_pca_ind(
  res.pca,
  geom.ind = "point",
  pointsize = 3
) +
  theme_classic()

# Graphique des variables
fviz_pca_var(
  res.pca,
  col.var = "contrib"
) +
  theme_classic()


```

```{r}
#| label: check-performance-models

modele_lm_2 <- lm(estimate ~ leaf_age * age_class, data = May_young_old)
modele_lm_1 <- lm(estimate ~ leaf_age, data = May_young_old)

AIC(modele_lm_1, modele_lm_2)
BIC(modele_lm_1, modele_lm_2)

anova(modele_lm_1, modele_lm_2)
summary(modele_lm_1)$adj.r.squared
summary(modele_lm_2)$adj.r.squared

```

```{r}

```

# Hot T50

```{r}
#| label: Create-joint-data-frame-HT50-traits

# create data frame with T50 estimate for Hot tolerance
united_columns_HT50_plante <- TTHot_wide_plant |> 
  filter(metric == "T50") |> 
  dplyr::select(site_name, species, month, age_class, plant_nr, leaf_age, estimate, se, lower, upper) |> # carefull in the metric we only want T50 
  unite(
    col = "new_id",
    site_name, species, month, age_class, plant_nr, leaf_age,
    sep = "_",
    remove = TRUE
  ) |> 
  na.omit()

# create a joint data frame with trais and T50
united_columns_joint_traits_HT50 <- left_join(united_columns_traits_plante, united_columns_HT50_plante, by = "new_id")

# String split back to columns the ID
joint_traits_HT50_wide <- united_columns_joint_traits_HT50 |> 
  separate_wider_delim(
    cols = new_id, 
    delim = "_", 
    names = c("site_name", "species", "month", "age_class", "plant_nr", "leaf_age")
  ) |> 
  mutate(
    # Check the data type of the new output columns and convert if necessary
    across(c(site_name, species, plant_nr, leaf_age), factor),
    # Relevel factors in natural order
    month = factor(month, levels = c("March", "May")),
    age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer", "Building", "Mature"))
  )
```

```{r}
#| label: Scale-joint-data-frame-HT50-traits

joint_traits_HT50_scaled <- joint_traits_HT50_wide |> 
  mutate(
    across(c(height_flower, height_vegetative, leaf_area, wet_mass, dry_mass, SLA, LDMC), ~ as.numeric(as.character(.x))),
    across(c(height_flower, height_vegetative, leaf_area, wet_mass, dry_mass, SLA, LDMC), ~ as.numeric(scale(.x)), .names = "{.col}_scaled")
  )

```

## March, May, NB, PIO, old

```{r}
#| label: Create-data-frame

joint_traits_HT50_NB_PIO_old <- joint_traits_HT50_wide |>
  filter(age_class %in% c("Newly Burnt", "Pioneer")) |>
  filter(leaf_age %in% c("old"))
```

```{r}
#| label: Create-scaled-data-frame

joint_traits_HT50_NB_PIO_old_scaled <- joint_traits_HT50_NB_PIO_old |> 
  mutate(
    across(c(height_flower, height_vegetative, leaf_area, wet_mass, dry_mass, SLA, LDMC), ~ as.numeric(as.character(.x))),
    across(c(height_flower, height_vegetative, leaf_area, wet_mass, dry_mass, SLA, LDMC), ~ as.numeric(scale(.x)), .names = "{.col}_scaled")
  )
```

```{r}
#| label: Correlation-between-traits

traits_data_corr <- joint_traits_HT50_NB_PIO_old |>
  dplyr::select(height_flower, height_vegetative, dry_mass, wet_mass, leaf_area, SLA, LDMC)

corr_matrix = round(cor(traits_data_corr, use = "complete.obs"), 2) #ignoring NA
ggcorrplot(corr_matrix, hc.order = TRUE, type = "lower", lab = TRUE)

# Cut-off value : 0.70 --> literature
# dry_mass and wet_mass : 0.85
# leaf_area and dry_mass : 0.76
# leaf_area and wet_mass : 0.77
# height_flower and heigth_vegetative : 0.95

# We removed wet and dry mass because they are correlated with leaf area and with each other, and they are already represented in the LDMC.
# Between vegetative and flowering height, we removed flowering height because the experiment focuses on leaves.

traits_data_2 <- data |>
  filter(proj_2025 == "Traits") |>
  dplyr::select(height_vegetative, leaf_area, SLA, LDMC)

corr_matrix = round(cor(traits_data_2, use = "complete.obs"), 2) #ignoring NA
ggcorrplot(corr_matrix, hc.order = TRUE, type = "lower", lab = TRUE)

# Four variables left for the model
```

```{r}
#| label: Data-frame

joint_traits_HT50_NB_PIO_old <- joint_traits_HT50_NB_PIO_old |>
  dplyr::select(-c(height_flower, wet_mass, dry_mass))

joint_traits_HT50_NB_PIO_old_scaled <- joint_traits_HT50_NB_PIO_old_scaled |>
  dplyr::select(-c(height_flower, wet_mass, dry_mass))
```

#### Trait-by-trait models

```{r}
#| label: SLA ~ age_class * month
#| fig-height: 10

mod_SLA  <- lm(SLA ~ age_class * month, data = joint_traits_HT50_NB_PIO_old)
performance::check_model(mod_SLA)

anova(mod_SLA)

# 1. Normality (Q-Q Plot): Valid. Most residuals follow the line well. 
#    Minor deviation at the upper tail (plants 77, 87, 89) but nothing critical.
# 2. Homoscedasticity: Valid. Residuals are evenly distributed across groups 
#    (the reference line remains flat and horizontal).
# 3. Collinearity: Low (VIF < 5 for all terms).
```

```{r}
#| label: LDMC ~ age_class * month
#| fig-height: 10
mod_LDMC <- lm(LDMC ~ age_class * month, data = joint_traits_HT50_NB_PIO_old)
performance::check_model(mod_LDMC)

anova(mod_LDMC)

# 1. Normality (Q-Q Plot): Valid. Residuals follow the line.
# 2. Homoscedasticity: Valid. Residuals show a stable distribution across groups
#    (the reference line is slightly curved but acceptable).
# 3. Collinearity: Low (VIF < 5 for all terms). 
```

```{r}
#| label: leaf_area ~ age_class * month
#| fig-height: 10
mod_LA   <- lm(leaf_area ~ age_class * month, data = joint_traits_HT50_NB_PIO_old)
performance::check_model(mod_LA)

anova(mod_LA)

# 1. Normality (Q-Q Plot): Valid. Residuals follow the line closely, with a very slight deviation at the right tail, which is acceptable.
# 2. Homoscedasticity: Valid. Uniform spread of residuals across fitted values 
#    (the reference line is straight and horizontal).
# 3. Collinearity: Low (VIF < 5). 
```

```{r}
#| label: height_vegetative ~ age_class * month
#| fig-height: 10
mod_hveg <- lm(height_vegetative ~ age_class * month, data = joint_traits_HT50_NB_PIO_old)
performance::check_model(mod_hveg)

anova(mod_hveg)

# 1. Normality (Q-Q Plot): Valid. Residuals follow the reference line.
# 2. Homoscedasticity: Valid. Residual distribution is stable (the slight upward 
#    curve at the right edge is minor and acceptable given the extreme F-value).
# 3. Collinearity: Low (VIF < 5). No issues.

# Note on F-value (1659.4): explain by the height difference between Newly Burnt (small resprouts) and Pioneer (tall bushes) 
```

**Summary of Univariate ANOVA Models on Functional Traits**

| Functional Trait | Age Class Effect (`age_class`) | Season Effect (`month`) | Interaction Effect (`Age x Month`) |
|:-----------------|:----------------:|:----------------:|:----------------:|
| SLA (Specific Leaf Area) | Ultra-significant (\*\*\*) | Ultra-significant (\*\*\*) | Non-significant |
| LDMC (Leaf Dry Matter Content) | Non-significant | Ultra-significant (\*\*\*) | Non-significant |
| Leaf Area | Ultra-significant (\*\*\*) | Marginal / Tendency (.) | Non-significant |
| Height (Vegetative) | Ultra-significant (\*\*\*) | Ultra-significant (\*\*\*) | Ultra-significant (\*\*\*) |

#### Final model

```{r}
#| label: Final-model
#| fig-height: 10

mod_final_additif <- lm(estimate ~ SLA_scaled + leaf_area_scaled + LDMC_scaled + height_vegetative_scaled + month, data = joint_traits_HT50_NB_PIO_old_scaled) 


mod_final_interaction <- lm(estimate ~ (SLA_scaled + leaf_area_scaled + LDMC_scaled + height_vegetative_scaled) * month, data = joint_traits_HT50_NB_PIO_old_scaled) 


AIC(mod_final_additif, mod_final_interaction)

# Le modèle avec interaction (mod_final_interaction) a un AIC plus bas de $8.66$ points par rapport au modèle additif.En statistiques (et plus particulièrement avec le critère de l'AIC), dès qu'une différence est supérieure à $2$ points, le modèle avec l'AIC le plus bas est considéré comme significativement meilleur. Une différence de près de $9$ points est donc un signal très fort.

performance::check_model(mod_final_interaction)

# 1. Normality (Q-Q Plot): Valid. Residuals follow the reference line tightly.
# 2. Collinearity (VIF): Valid. Low to moderate VIF (< 6) across all terms, 
#    which is exceptionally clean for a complex interaction model.
# 3. Homoscedasticity: Acceptable. Minor hump in residual spread for intermediate 
#    fitted values, but well within normal limits for ecological field data.
# 4. Influential Observations: No overly influential outliers detected.

summary(mod_final_interaction)
```

```{r}
#| label: effect-of-traits-on-T50
#| warning: false

class_colors <- c("Newly Burnt" = "#D95F02", "Pioneer" = "#1B7C7A")
class_shapes <- c("Newly Burnt" = 16, "Pioneer" = 17)

common_style <- list(
  facet_wrap(~ month),
  scale_color_manual(values = class_colors, name = "Age Class"),
  scale_shape_manual(values = class_shapes, name = "Age Class"),
  theme_classic(base_size = 12),
  theme(
    panel.background = element_rect(fill = "#FAFAFA", color = NA),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 13, color = "#333333"),
    legend.position = "bottom" 
  )
)

p_sla <- ggplot(joint_traits_HT50_NB_PIO_old_scaled, aes(x = SLA, y = estimate)) +
  geom_smooth(method = "lm", color = "#222222", fill = "grey85", linetype = "solid", size = 0.7, alpha = 0.5) +
  geom_point(aes(color = age_class, shape = age_class), size = 2.5, alpha = 0.8) + 
  labs(x = "SLA (mm²/g)", y = "T50 (°C)") +
  common_style

p_ldmc <- ggplot(joint_traits_HT50_NB_PIO_old_scaled, aes(x = LDMC, y = estimate)) +
  geom_smooth(method = "lm", color = "#222222", fill = "grey85", linetype = "solid", size = 0.7, alpha = 0.5) +
  geom_point(aes(color = age_class, shape = age_class), size = 2.5, alpha = 0.8) + 
  labs(x = "LDMC", y = "T50 (°C)") +
  common_style

p_area <- ggplot(joint_traits_HT50_NB_PIO_old_scaled, aes(x = leaf_area, y = estimate)) +
  geom_smooth(method = "lm", color = "#222222", fill = "grey85", linetype = "solid", size = 0.7, alpha = 0.5) +
  geom_point(aes(color = age_class, shape = age_class), size = 2.5, alpha = 0.8) + 
  labs(x = "Leaf Area (mm²)", y = "T50 (°C)") +
  common_style


p_height <- ggplot(joint_traits_HT50_NB_PIO_old_scaled, aes(x = height_vegetative, y = estimate)) +
  geom_smooth(method = "lm", color = "#222222", fill = "grey85", linetype = "solid", size = 0.7, alpha = 0.5) +
  geom_point(aes(color = age_class, shape = age_class), size = 2.5, alpha = 0.8) + 
  labs(x = "Height vegetative (cm)", y = "T50 (°C)") +
  common_style

print(p_sla)
print(p_ldmc)
print(p_area)
print(p_height)

```

```{r}
#| label: Correlation-matrix-of-functional-traits-and-T50
# #| warning: false
# 
# # if(!require(GGally)) install.packages("GGally")
# # if(!require(ggplot2)) install.packages("ggplot2")
# 
# library(GGally)
# 
# 
# traits_HT50 <- joint_traits_HT50_NB_PIO_old[, c("estimate", "SLA", "height_vegetative", "leaf_area", "LDMC")]
# 
# 
# ggpairs(
#   traits_HT50,
#   upper = list(continuous = wrap("cor", method = "pearson", size = 4.5)),
#   diag = list(continuous = wrap("densityDiag", fill = "lightgray", alpha = 0.5)),
#   lower = list(continuous = wrap("smooth", method = "lm", color = "blue", alpha = 0.3, size = 0.5))
# ) +
# theme_bw() + 
# labs(title = "Correlation matrix of functional traits and T50") +
# theme(
#   strip.text = element_text(size = 11, face = "bold"), 
#   plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
# )
```

```{r}
#| label: prediction-effect
#| warning: false


library(ggeffects)
library(patchwork)

# Prédictions
pred_SLA <- ggpredict(mod_final_interaction,
                      terms = c("SLA_scaled", "month"))

pred_LDMC <- ggpredict(mod_final_interaction,
                       terms = c("LDMC_scaled", "month"))

pred_LA <- ggpredict(mod_final_interaction,
                     terms = c("leaf_area_scaled", "month"))

pred_H <- ggpredict(mod_final_interaction,
                    terms = c("height_vegetative_scaled", "month"))

# Fonction pour tracer les prédictions
plot_pred <- function(pred, xlab){

  ggplot(pred,
         aes(x = x,
             y = predicted,
             colour = group,
             fill = group)) +

    geom_line(linewidth = 1.2) +

    geom_ribbon(aes(ymin = conf.low,
                    ymax = conf.high),
                alpha = 0.2,
                colour = NA) +

    labs(x = xlab,
         y = "T50 of Fv/Fm",
         colour = "Month",
         fill = "Month") +

    theme_classic(base_size = 14)

}

# Les quatre graphiques
p1 <- plot_pred(pred_SLA, "SLA")
p2 <- plot_pred(pred_LDMC, "LDMC")
p3 <- plot_pred(pred_LA, "Leaf area")
p4 <- plot_pred(pred_H, "Height")

# Assemblage
(p1 | p2) /
(p3 | p4)
```

```{r}
#| label: prediction-effect
#| warning: false
library(broom)


coef_mod <- tidy(mod_final_interaction, conf.int = TRUE) |>
  filter(term != "(Intercept)")

coef_mod <- coef_mod |>
  mutate(
    term = dplyr::recode(term,
      "SLA_scaled" = "SLA",
      "leaf_area_scaled" = "Leaf area",
      "LDMC_scaled" = "LDMC",
      "height_vegetative_scaled" = "Vegetative height",

      "monthMay" = "Month: May",

      "SLA_scaled:monthMay" = "SLA × May",
      "leaf_area_scaled:monthMay" = "Leaf area × May",
      "LDMC_scaled:monthMay" = "LDMC × May",
      "height_vegetative_scaled:monthMay" = "Vegetative height × May"
    ),

    type = ifelse(grepl("×", term), "Interaction", "Main effect")
  )

coef_mod$term <- factor(
  coef_mod$term,
  levels = c(
    "Month: May",
    "SLA",
    "Leaf area",
    "LDMC",
    "Vegetative height",
    "SLA × May",
    "Leaf area × May",
    "LDMC × May",
    "Vegetative height × May"
  )
)

# Forest plot
ggplot(coef_mod,
       aes(x = estimate,
           y = term)) +


  geom_vline(xintercept = 0,
             linetype = "dashed",
             colour = "grey50") +


  geom_hline(yintercept = c(1.5, 5.5),
             colour = "grey75",
             linewidth = 0.6) +


  geom_errorbar(
    aes(xmin = conf.low,
        xmax = conf.high),
    width = 0.15,
    linewidth = 0.8
  ) +


  geom_point(size = 3.5) +

  labs(
    x = expression(Standardized~coefficient~(beta)),
    y = NULL
  ) +

  theme_classic(base_size = 15) +

  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14)
  )
```

# New: Effect of leaf_age_2 and age_class on traits (pca dim)

## Import dataset

```{r}
#| label: Import-dataset

pca_dim <- readRDS("C:/Users/poula/Documents/M2_PAMEQ/Stage M2/Analyse statistique/thermotolerance/data/clean_data/pca_dim.rds")

```

## Models

```{r}
#| label: models
#| fig-height: 10

# Model for PCA 1 (Dim1)
model_pca1 <- lm(PCA1_score ~ age_class * leaf_age_2, data = pca_dim)
performance::check_model(model_pca1)
summary(model_pca1)
anova(model_pca1)

# Model for PCA 2 (Dim2)
model_pca2 <- lm(PCA2_score ~ age_class * leaf_age_2, data = pca_dim)
performance::check_model(model_pca2)
summary(model_pca2)
anova(model_pca2)

```
