---
title: "DURIN 2026 Seasonal Thermal Tolerance Calluna data cleaning"
author: "Sonya R Geange and Gaëlle Poulat"
format: html
editor: visual
---

# Library

```{r}
#| label: loading-libraries
#| message: false 

# installing / loading packages

library(readxl) # Enables excel file import
library(tidyverse) # Contains many helpful packages, i.e. dplyr etc.
library(janitor) # Has tools for data cleaning 
library(lubridate) # Date and time formatting
library(readr) # Data importation functions
library(stringr) # String and character text
library(future) # Using severals cores at the same time
library(ggcorrplot)
library(emmeans)
library(car) # for the Anova
library(dplyr)
```

# Import Data

```{r}
#| label: function-clean-trait-dataset-to-match-DURIN-formatting

clean_sheet_data <- function(file_path, sheet_name) {
  
  # Importing raw data of trait and fluorescence
  df <- read_excel(file_path, sheet = sheet_name, na = c("", "NA"))
  
  # Clean trait dataset to match DURIN formatting
  df <- df |> 
    # New name then old name of column, including snake case ('_' and lower letters)
    dplyr::rename(
      site_name = siteID,
      age_class = ageClass,
      collection_time = CollTime,
      paired_with = PairedWith,
      temp_exp = TempExp,
      temp_exp_time = TempExpTime,
      temp_exp_time_unit = TempExpTimeUnit,
      fv_fm_initial = 'Fv/Fm_Initial',
      fv_fm_hr = 'Fv/Fm_Hr',
      fv_fm_final = 'Fv/Fm_Final',
      height_flower = 'height(s)_flower',
      height_vegetative = 'height(s)_top',
      treatment_date = TreatmentDate,
      treatment_time = TreatmentTime
    ) |> 
    # using slice() to remove the first three row of example data (ALWAYS CHECK!)
    slice(-(1:3)) |> 
    
    # Changing classes of all variables that should be in class factor
    mutate(across(c(
      site_name, species, proj_2025, project, age_class, plant_nr, leaf_age, 
      sample, paired_with, calluna_shoot_type, temp_exp, temp_exp_time, 
      temp_exp_time_unit, fv_fm_hr, scanned
    ), as.factor)) |> 
    
    # Changing classes of variables that should be in class dbl/numeric
    # (suppressWarnings évite le message 'NAs introduits lors de la conversion')
    mutate(across(c(
      fv_fm_initial, fv_fm_final, wet_mass_g, leaf_thickness_1_mm, 
      leaf_thickness_2_mm, leaf_thickness_3_mm, dry_mass_g, height_flower, height_vegetative
    ), ~ suppressWarnings(as.numeric(.)))) |> 
    
    # Changing classes of all variables that should be in class integers
    mutate(across(c(bulk_nr_leaves, leaf_nr), ~ suppressWarnings(as.integer(.)))) |> 
    
    # Order relevant variables (useful for plotting and order based analyses)
    # age_class sorted from youngest to oldest
    mutate(age_class = factor(age_class, levels = c("Newly Burnt", "Pioneer", "Building", "Mature"))) |> 
    
    # Extraction sécurisée de l'heure et création de la date
    mutate(
      time_clean = format(parse_date_time(collection_time, orders = c("HMS", "HM", "Ymd HMS")), "%H:%M"),
      date = ymd_hm(paste(year, month, day, time_clean)), 
      .after = 1
    ) |> 
    select(-c(day, year, collection_time, time_clean)) |> 
    
    # Create a unique plant_ID given age_class, month, and plant_nr
    mutate(unique_plantID = paste(age_class, month, plant_nr, sep = "_"))
  
  # Correction spécifique au code barre si la colonne existe
  if ("envelope_ID" %in% names(df)) {
    df <- df |> 
      mutate(envelope_ID = case_when(
        grepl("ADDX1933", envelope_ID) ~ "ADX1933",
        grepl("AGS", envelope_ID) ~ "AGS4838",
        TRUE ~ envelope_ID
      ))
  }
  
  return(df)
}


```

```{r}
#| label: clean-trait-dataset-to-match-DURIN-formatting-function

path <- "C:/Users/poula/Documents/M2_PAMEQ/Stage M2/Analyse statistique/thermotolerance/data/raw_data/Seasonal Calluna Thermal Tolerance and Plant Functional Traits.xlsx"

raw_data_mar  <- clean_sheet_data(path, sheet_name = "cv_mar26")
raw_data_may  <- clean_sheet_data(path, sheet_name = "cv_may26")
raw_data_july <- clean_sheet_data(path, sheet_name = "cv_july26")

# NOTE: Fixes date parsing warning (line 684) — handled below in the 'adjusting-weird-values-in-the-data' section.
```

## Merge month dataset

```{r}
#| label: Merge-March-date-frame-and-May-data-frame

raw_data <- bind_rows(raw_data_mar, raw_data_may, raw_data_july)

raw_data <- raw_data |> 
  mutate(
    month = factor(month, levels = c("March", "May", "July")),
    age_class = factor(age_class, 
                       levels = c("Newly Burnt", "Pioneer", "Building", "Mature"),)
  )
```

## Leaf area dataset

```{r}
#| label: import-leaf-area-dataset-tidy-and-match-to-trait-dataset


# Change this to your local repository where the four excel output files are stored
leaf_area_dir <- "C:/Users/poula/Documents/M2_PAMEQ/Stage M2/scans/cv_output/"

# Import leaves from the directory where the files are saved, only .csv files, return the file paths (not just file names)
leaf_area_raw <- list.files(leaf_area_dir, pattern = "\\.csv$", full.names = TRUE) |> 
  # Give each element the name of the source file (remove directory path)
  (\(x) setNames(x, basename(x)))() |>
  # Read each .csv file into R, combine into a single dataframe
  # Add an ID column, indicating the source file
  purrr::map_df(~ readr::read_csv(.x, show_col_types = FALSE), .id = "source_file")

#leaf_area_raw has only 2 columnes and we have to separate them
leaf_area_raw <- leaf_area_raw |> 
  separate(
    col =  `number;ID;n;leaf_area`,          # The column to separate
    into = c("number", "ID", "n", "leaf_area"), # Names for the new columns
    sep = ";",            # The delimiter to use for splitting
  ) |> 
  #remove the column number
  select(-c(number))

# Clean and summarize files to get leaf area per envelopeID
leaf_area_raw <- leaf_area_raw |> 
  dplyr::select(-dplyr::matches("^Unnamed")) |>   # drop index col if present
  dplyr::mutate(
    # Create new column names, and/or update the datatype for a given column
    envelope_ID = str_trim(as.character(ID)),
    leaf_area_scanned_bulk_nr = as.numeric(n),
    leaf_area_sum = as.numeric(leaf_area)
  ) |> 
  # remove old column, comments is deleted as should be in the trait datasheet area check tab
  select(-c(ID, n,leaf_area))




# Choose correct value per envelope_ID based on the source file, i.e. edited or not
leaf_area_clean <- leaf_area_raw |> 
  # Use the phrase 'edited_images' in the file name to prioritize
  mutate(priority = if_else(str_detect(source_file, "edited_images"), 1, 2)) |>   
  arrange(envelope_ID, priority) |>    # ordering so preferred rows come first
  distinct(envelope_ID, .keep_all = TRUE) |>   # keeping first row per ID
  select(-priority) # removing helper column

#ok! we have 165 enveloppes

# Join the clean leaf area db to the main trait db by the envelope ID
data <- raw_data |> 
  dplyr::mutate(envelope_ID = stringr::str_trim(as.character(envelope_ID))) |> 
  # Left join, keep values from x, and only matching rows in y (leaf_area_clean)
  dplyr::left_join(leaf_area_clean, by = "envelope_ID")

```

# Data Cleaning Entire Dataframe

Above was cleaning of column names, formatting of variables etc. Now we need to identify any mistakes in the data entry, outliers etc.

```{r}
#| label: standardize-comments-and-remarks

# Find unique comments in the dataset
# unique(data$remark)

# Multiple comment strings in the column (EXAMPLE ONLY AT THE MOMENT!)
## first observation : some REMARKS are the same but with capital letters for example

data <- data |> 
  mutate(
    # 1. Basic cleaning
    remark = tolower(remark),
    remark = gsub("\\s*;\\s*", "; ", remark),
    remark = str_trim(remark),
    
    # 2. Pattern-based substitutions
    remark = gsub("\\b[^;]*((leaf|leaves)[^;]*(sellotape|glue)|glue)[^;]*", "leaf stuck to tape", remark),
    remark = gsub("only\\s+(3 leaves at scanning)", "\\1", remark),
    
    # Term-specific corrections
    remark = gsub("\\bfluorescence\\b|\\bfluo\\b", "fv/fm", remark),
    remark = gsub("\\bdyr mass\\b", "dry mass", remark),
    remark = gsub("paper towel was a little dry", "paper towel was little dry", remark),
    remark = gsub("no flowers\\b", "no flower", remark),
    
    # Append 'at fv/fm' to '3 leaves' or '4 small leaves' 
    # (only if not already followed by 'at' or 'after')
    remark = gsub("\\b(3 leaves|4 small leaves)\\b(?!\\s+(at|after))", "\\1 at fv/fm", remark, perl = TRUE),
    
    # Clean up trailing commas and trailing semicolons
    remark = gsub(",\\s*", "; ", remark),
    remark = gsub(";\\s*$", "", remark),

    # 3. Extract fragment containing 'brown'
    scanning = str_extract(remark, "[^;]*brown[^;]*") |> str_trim(),

    # 4. Specific conditional corrections for scanning
    scanning = case_when(
      month == "March" & proj_2025 == "Traits" & age_class == "Building"   & plant_nr == "1" & leaf_nr == "20" ~ "mostly brown leaves",
      month == "March" & proj_2025 == "Traits" & age_class == "Building"   & plant_nr == "6" & leaf_nr == "18" ~ "brown leaves",
      month == "March" & proj_2025 == "Traits" & age_class == "Newly Burnt" & plant_nr == "7"                  ~ "partly brown leaves",
      month == "March" & proj_2025 == "Traits" & age_class == "Building"   & plant_nr == "8"                  ~ "brown leaves",
      TRUE ~ scanning
    ),

    # 5. Global overrides
    remark = case_when(
      grepl("water", remark)     ~ "liquid water in bag",
      grepl("fridge", remark)    ~ "not removed from fridge, no balls",
      grepl("harvested", remark) & !grepl("harvested after 24h", remark) ~ "no leaf harvested",
      
      # Specific element reordering
      remark == "3 leaves at wet mass; 3 leaves at thickness; mostly brown leaves" ~ 
        "3 leaves at scanning; mostly brown leaves; 3 leaves at thickness",
      envelope_ID == "APL2797" ~ "3 leaves at scanning; 2 leaves at thickness; 2 leaves at dry mass",
        
      # Specific note update with sample ID
      remark == "wet mass only 4dp, 0.1069" ~ 
        "wet mass only 4dp, 0.1069 for AUX097; value changed to 0.01069",
        
      TRUE ~ remark
    )
  )

unique(data$remark)

```

```{r}
#| label: adjusting-scanned_bulk_nr

# While the bulk number from the envelope represents the initial sample size, sometimes leaves are lost or gained during the trait sampling process. Need to check the particule count from the leaf area scans to adjust the bulk number leaves. This needs to be done before calculating per leaf trait values, which can be used for metrics like SLA and LDMC.

# BULK LEAF COUNT ADJUSTMENTS:
# - Pre-processing (copies of original images): Artifacts (dust/leftover fragments) erased and stuck leaves separated using Paint prior to re-running ImageJ.
# - Script-level fixes: Subtracts partial leaf fragments to reflect actual scanned leaf numbers per envelope.
# - Validation: All non-standard bulk counts (!= 4) left unmodified were manually cross-checked.

data <- data |>
  mutate(
    leaf_area_scanned_bulk_nr = case_when(
      # March
      # Partial leaf fragment; unknown parent leaf
      envelope_ID %in% c("ABW4300", "ABX8228", "ABY3458", "AES5198", "AEY6133","AFD4235", "AFE4708", "AFF6622", "AFG3152", "AFH2775", "DNH5015", "DOF6452", "ACL6886", "ACW2990", "ADC7047", "ADS2507", "AGR6304") ~ 4,
   
      # May
      # Partial leaf fragment; unknown parent leaf
      envelope_ID %in% c("AIR6318", "AMI3269", "ANP7023", "AQI1129", "AQT4821") ~ 4,

      # July
      # Partial leaf fragment; unknown parent leaf
      envelope_ID %in% c("AUL1455", "BAX3476", "BAY7128", "BBZ0443", "BCA1604") ~ 4,
      
      # Partial leaf fragment; known parent leaf
      envelope_ID == "BCH3466" ~ 4,
      
      TRUE ~ leaf_area_scanned_bulk_nr
    )
  )



```

```{r}
#| label: adjusting-weird-values-in-the-data

## March
data <- data |> 
  mutate(leaf_thickness_3_mm = case_when(
    grepl("AGT0463", envelope_ID) ~ 0.249,
    TRUE ~ leaf_thickness_3_mm
  )) |>   #BUI 8 : 0.588 in data frame ; 0.249 on the envelope AGT0463
  mutate(month = case_when(
    ymd_hms(date) == ymd_hms("2026-01-10 08:00:00") ~ "March",
    TRUE ~ month
  )) |>  #wrong month for PIO1 - leaf15
  mutate(date = case_when(
    ymd_hms(date) == ymd_hms("2026-01-10 08:00:00") & month == "March" ~ ymd_hms("2026-03-10 08:00:00"),
    TRUE ~ date
  )) |>  #wrong month for PIO1 - leaf15
  mutate(wet_mass_g = case_when(
    grepl(NA, wet_mass_g) ~ 0.01063,
    TRUE ~ wet_mass_g
  )) #NB 7 : no value in data frame ; 0.01063 on the envelope AFN1989


## May 
data <- data |> 
## wet mass
  filter(envelope_ID != "AJD4636" | is.na(envelope_ID))|>  #NB 4 : 0.05708 in data frame ; 0.05708 on the envelope AJD4636 and 4 leaves on the scanns. Excluded sample: presence of non-target branching leaf causing artificially high wet mass.
  mutate(wet_mass_g = case_when(
    grepl("AUX0977", envelope_ID) ~ 0.01069,
    TRUE ~ wet_mass_g
  )) |>   #PIO 7 : 0.10690 in data frame ; 0.1069 on enveloppe AUX097--> pb with 5 decimals scale
  mutate(wet_mass_g = case_when(
    grepl("APP8997", envelope_ID) ~ 0.01449,
    TRUE ~ wet_mass_g
  )) |>   #MAT 2 : 0.14490 in data frame ; 0.01449 on the envelope APP8997
  mutate(dry_mass_g = case_when(
    grepl("AOM6256", envelope_ID) ~ 0.00590,
    TRUE ~ dry_mass_g
  )) |>   #MAT 8 : 0.0590 in data frame ; 0.00590 on the enveloppe AOM6256
## wrong month  
  mutate(month = case_when(
    month == "March" & proj_2025 == "TT Hot" & age_class == "Pioneer" & plant_nr == "6" & leaf_age == "old" & temp_exp == 38 & fv_fm_final == 0.12 ~ "May",
    month == "March" & proj_2025 == "TT Hot" & age_class == "Pioneer" & plant_nr == "6" & leaf_age == "young" & temp_exp == 38 & fv_fm_final == 0.69 ~ "May",
    TRUE ~ month
  )) |> 
  mutate(date = case_when(
    is.na(date) | date %in% c("na", "NA") ~ "2026-05-19 08:11:00",
    TRUE ~ as.character(date)
  )) |> 
  mutate(date = ymd_hms(date)) |> 
#LDMC
  filter(envelope_ID != "APL2797" | is.na(envelope_ID)) |> # BUI 2 (APL2797) : 1.720, too high. 4 leaves at wet mass (0.0043/leaf) and 2 leaves at dry mass (0.00074).Excluded sample.
# NB 8 young excluded sample for traits: presence of non-target branching leaf causing artificially high measure. For fv:fm, no branching leaf, kept in the data.
  filter(
    !(month %in% "May" & 
      proj_2025 %in% "Traits" & 
      age_class %in% "Newly Burnt" & 
      plant_nr %in% "8" & 
      leaf_age %in% "young")
  )


## July 
data <- data |> 
## wet mass
  mutate(wet_mass_g = case_when(
    grepl("AWR0569", envelope_ID) ~ 0.01261,
    TRUE ~ wet_mass_g
  ))  #PIO 3 : 0.001261 in the data frame ; 0.01261 on envelope AWR0569
#Excluded samples: BUI 1 and 2 were measured twice: immediately and after 24 hours du to the presence of non-target branching leave in the first measurement. A  difference is observed between the two time points, with the 24-hour measurements being more consistent with the other buildings. Consequently, only the 24-hour measurements for BUI 1 and 2 were retained in the dataset. --> done in the July Building 1 and 2 issue section.
```

# Calculate values per leaf trait

```{r}
#| label: adjusting-bulk_nr-for-each-trait

# While the bulk number from the envelope represents the initial sample size, sometimes leaves are lost or gained during the trait sampling process. Need to check comments and the particule count from the leaf area scans to adjust the bulk number leaves for each trait. This needs to be done before calculating per leaf trait values, which can be used for metrics like SLA and LDMC.

# Relies on cleaned remarks column from step above

# Note, this is only for wet mass, leaf area, dry mass.
# For fluorescence this is per sample in the clip
# For thickness, each measurement (1-3) needs to be checked first, then averaged in cae one of the replicate samples is an outlier and will skew the average thickness


# Default column is the intiial bulk_nr, initial values for following trait derived from prior trait + edits via comments

data$bulk_nr_wet_mass <- data$bulk_nr_leaves


# Modify based on the wet mass comments
data <- data |> 
  mutate(bulk_nr_wet_mass = case_when(
    grepl("3 leaves at wet mass", remark) ~ 3,
    grepl("2 leaves at wet mass", remark) ~ 2,
    grepl("5 leaves at wet mass", remark) ~ 5,
    .default = bulk_nr_wet_mass    # Keep original bulk nr value if no comment is applied to row
  ))

# Create leaf area, rename the particle counts from scanner. More accurate then comments for values less than 4. However, sometimes leaves break into multiple parts, so the particle count may be higher than the leaf count artificially.
data <- data |> 
  dplyr::rename(bulk_nr_leaf_area = leaf_area_scanned_bulk_nr)

# Modify based on the wet mass comments
data <- data |> 
  mutate(bulk_nr_leaf_area = case_when(
    # Checked against images, indeed 5 leaves (not broken)
    grepl("2 leaves at scanning", remark) ~ 2,
    grepl("5 leaves at scanning", remark) ~ 5,
    grepl("3 leaves at scanning", remark) ~ 3,
    .default = bulk_nr_leaf_area    # Keep original bulk nr value if no comment is applied to row
  ))

# Check count of leaf numbers by the remark column to help identify issues
# We have 5, 6, 7 and 8's
table(data$remark, data$bulk_nr_leaf_area)

# Find rows where leaf number for area is higher than leaf number for wet mass
# Need to go and check 1) leaf scans, 2) envelopes to see if extra particles are in envelope, 3) remarks column, 4) the excel tab on leaf area checks
bulk_nr_issues <- data |> 
  filter(bulk_nr_leaf_area > bulk_nr_wet_mass) |> 
  select(envelope_ID, remark, bulk_nr_leaves, bulk_nr_wet_mass, bulk_nr_leaf_area)

# All other envelopes have correct number based on the wet mass column in the images, so correct back to the wet mass values

# Modify based on the comments, checking images etc. 
data <- data |> 
  mutate(bulk_nr_leaf_area = case_when(
    grepl("2 leaves at scanning", remark) ~ 2,
    grepl("5 leaves at scanning", remark) ~ 5,
    grepl("3 leaves at scanning", remark) ~ 3,
    .default = bulk_nr_wet_mass    # Keep original bulk nr value if no comment is applied to row
  ))


# Note, for thickness the value is note correct by bulk leaf number, however, if there are fewer than three leaves, there may only be one or two leaf thickness measurements to average across. 
data$bulk_nr_thickness <- data$bulk_nr_leaf_area

# Check how many envelopes have fewer than three thickness measurements
bulk_nr_issues_thickness <-  data |> 
  filter(rowSums(across(starts_with("leaf_thickness_"), ~ . > 0)) < 3)


# Modify based on the thickness comments
data <- data |> 
  mutate(bulk_nr_thickness = case_when(
    grepl("2 leaves at thickness", remark) ~ 2,
    grepl("3 leaves at thickness", remark) ~ 3,
    grepl("5 leaves at thickness", remark) ~ 5,
    .default = bulk_nr_thickness    # Keep original bulk nr value if no comment is applied to row
  ))


# Create dry mass bulk nr based on thickness
data$bulk_nr_dry_mass <- data$bulk_nr_thickness

# Modify based on the dry mass comments
data <- data |> 
  mutate(bulk_nr_dry_mass = case_when(
    grepl("2 leaves at dry mass", remark) ~ 2,
    grepl("3 leaves at dry mass", remark) ~ 3,
    grepl("5 leaves at dry mass", remark) ~ 5,
    .default = bulk_nr_dry_mass    # Keep original bulk nr value if no comment is applied to row
  ))
```

```{r}
#| label: calculate-per-leaf-values-for-each-trait

data <- data |>  
  mutate(
    wet_mass = wet_mass_g / bulk_nr_wet_mass,
    leaf_area = leaf_area_sum / bulk_nr_leaf_area,
    dry_mass = dry_mass_g / bulk_nr_dry_mass
  ) |> 
  # Round the decimal places
  mutate(
    wet_mass = round(wet_mass, digits = 5),
    leaf_area = round(leaf_area, digits = 3),
    dry_mass = round(dry_mass, digits =  5)
  )
```

# Check trait outliers

```{r}
#| label: check-plant-height-(vegetative)-outliers
#| warning: false

# making a scatterplot for vegetative plant height per plant
(data |> 
         filter(proj_2025 == "Traits") |>  
       ggplot(aes(x = factor(plant_nr), y = height_vegetative, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Plant Height - Vegetative (cm)",
    title = "Mean Vegetative Plant Height (cm) by plant, age class and month"
  ))


# making a scatterplot for flowering plant height per plant
(data |> 
         filter(proj_2025 == "Traits") |>  
       ggplot(aes(x = factor(plant_nr), y = height_flower, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Plant Height - Flowering (cm)",
    title = "Mean Flowering Plant Height (cm) by plant, age class and month"
  ))


# Flowering vs vegetative plant height
# Flowering height should probably be taller than vegetative height
# So above the 1:1 line

ggplot(data |>   
      filter(proj_2025 == "Traits"),
      aes(x = height_vegetative, y = height_flower)) +
  
  # scatter points
  geom_point(aes(colour = factor(plant_nr), shape = factor(age_class)),
             size = 2, alpha = 0.7) +
  
  # 1:1 reference line
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey50") +
  
  # fitted regression line
  geom_smooth(method = "lm", se = TRUE, colour = "blue") +
  
  # optional: label
  labs(
    x = "Vegetative Height (cm)",
    y = "Flowering Height (cm)",
    title = "Flowering vs Vegetative Plant Height",
    subtitle = "Dashed line = 1:1 reference, blue line = fitted linear trend",
    colour = "plant_nr"
  ) +
  
  theme_bw()

# NOTE: 
# March: NB 7 without flowering height
# July: no flowering height

```

```{r}
#| label: check-wet-mass-outliers-(per-leaf)
#| warning: false

# making a scatterplot for wet mass mean for trait leaves, plant_nr on x-axis and weight in g on y-axis

data |>  
         filter(proj_2025 == "Traits") |> 
       ggplot(aes(x = factor(plant_nr), y = wet_mass, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Wet mass (g)",
    title = "Mean wet mass per leaf by plant, age class and month"
  )

# If needed data modifications are documented in the adjusting-weird-values-in-the-data section.

## March
# wet mass ok exept for :
#           PIO 5 : 0.02825 in data frame ; 0.02825 on the envelope ACP1838
#           NB 7 : no value in data frame ; 0.01063 on the envelope AFN1989

## May
# wet mass ok except for :
#           PIO 7 : 0.10690 in data frame ; 0.1069 on envelope AUX097 --> pb with 5 decimal scale
#           MAT 2 : 0.14490 in data frame ; 0.01449 on the envelope APP8997
#           NB 4 : 0.05708 in data frame ; 0.05708 on the envelope AJD4636 and 4 leaves on scann.Excluded sample: presence of non-target branching leaf causing artificially high wet mass.
#           NB 8 young : 0.04181 in data frame ; 0.04181 on the envelope AMC1987 and 4 leaves on scann.
#           NB 8 young : 0.03912 in data frame ; 0.03912 on the envelope AMD2883 and 4 leaves on scann.Excluded sample: presence of non-target branching leaf causing artificially high wet mass.
#           NB 8 young : 0.05026 in data frame ; 0.05026 on the envelope AME1094 and 4 leaves on scann.Excluded sample: presence of non-target branching leaf causing artificially high wet mass.
#           NB 8 young : 0.03936 in data frame ; 0.03936 on the envelope AMF9314 and 4 leaves on scann.Excluded sample: presence of non-target branching leaf causing artificially high wet mass.
#           NB 8 young : Excluded sample: presence of non-target branching leaf causing artificially high wet mass.

## July
# wet mass ok except for :
#           PIO 3 : 0.001261 in the data frame ; 0.01261 on envelope AWR0569
```

```{r}
#| label: check-leaf-area-outliers-(per-leaf)
#| warning: false

# making scatterpolt for leaf area mean for trait leaves, plant_nr on x-axis and mean leaf area on y-axis. by plant nr, age class, month 
ggplot(data |>  filter(proj_2025 == "Traits"),
       aes(x = factor(plant_nr), y = leaf_area, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf area (mm2)",
    title = "Leaf area by plant, age class and month"
  )

# Check source file issues
ggplot(data |>  filter(proj_2025 == "Traits"),
       aes(x = factor(plant_nr), y = leaf_area, colour = factor(source_file))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf area (mm2)",
    title = "Leaf area by plant, age class and month"
  )

# Check source file issues
ggplot(data |>  filter(proj_2025 == "Traits"),
       aes(x = factor(plant_nr), y = leaf_area, colour = factor(source_file))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ source_file) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(
    x = "Plant number",
    y = "Leaf area (mm2)",
    title = "Leaf area by plant, age class and month"
  )

# bigger leaves for NB and smaller for the others
# for PIO : leaf area small edited image bigger than non edited one
# small edited images seem often to be plant 5 to 8 and large one plante 1 to 4
```

```{r}
#| label: check-leaf-thickness-outliers-(per-measurement-1-3-replicates)
#| warning: false
# histogram for leaf thickness distribution for trait leaves, not mean, rather seperate reps

data_long_thickness <- data |> 
  filter(proj_2025 == "Traits") |> 
  pivot_longer( # to get the seperate columns reshaped
    cols = c(leaf_thickness_1_mm, leaf_thickness_2_mm, leaf_thickness_3_mm),
    names_to = "replicate",
    values_to = "leaf_thickness_mm"
  )

ggplot(data_long_thickness, #plotting drom long format
       aes(x = factor(plant_nr), y = leaf_thickness_mm, colour = factor(leaf_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf thickness (mm)",
    title = "Leaf thickness replicates by plant, age class and month"
  )


ggplot(data_long_thickness, #plotting drom long format
       aes(x = factor(plant_nr),
           y = leaf_thickness_mm,
           colour = factor(leaf_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(leaf_nr ~ age_class) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf thickness (mm)",
    title = "Leaf thickness replicates by plant, age class and month"
  )

# If needed data modifications are documented in the adjusting-weird-values-in-the-data section.

## March
#         BUI 8 : 0.588 in data frame ; 0.249 on the envelope AGT0463

## May, high value for :
#         NB 4 - 16 young : 0.857 (thickness 1) in the data frame ; 0.857 on the enveloppe AJA9484
#         NB 4 - 17 young : 0.778 (thickness 1) in the data frame ; 0.778 on the enveloppe AJB0231
#         NB 4 - 18 young : 0.751 (thickness 1) in the data frame ; 0.751 on the enveloppe AJC0356
```

```{r}
#| label: check-dry-mass-outliers-(per-leaf)
#| warning: false

# making a scatterplot for dry mass mean for trait leaves, plant_nr on x-axis and weight in g on y-axis
ggplot(data |> filter(proj_2025 == "Traits"),
       aes(x = factor(plant_nr), y = dry_mass, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Dry mass (g)",
    title = "Mean dry mass per leaf by plant, age class and month"
  )

# If needed data modifications are documented in the adjusting-weird-values-in-the-data section.

```

```{r}
#| label: wet-mass-vs-dry-mass
#| warning: false

ggplot(data |>  filter(proj_2025 == "Traits"),
       aes(x = wet_mass_g, y = dry_mass_g)) +
  
  # scatter points
  geom_point(aes(colour = factor(plant_nr)),
             size = 2, alpha = 0.7) +
  
  # 1:1 reference line
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey50") +
  
  # fitted regression line
  geom_smooth(method = "lm", se = TRUE, colour = "blue") +
  
  # optional: label
  labs(
    x = "Wet mass per leaf (g)",
    y = "Dry mass per leaf (g)",
    title = "Dry mass vs wet mass per leaf",
    subtitle = "Dashed line = 1:1 reference, blue line = fitted linear trend",
    colour = "plant_nr"
  ) +
  
  theme_bw()

# according to the graph, dry mass always smaller than wet mass --> ok
# most of the plant have similar dry and wet mass
# relationship between dry and wet mass seem to be linear
# 2 low values for dry mass plant 6 for BUI and MAT (according to the graph just above) --> explain by the fact that the dry mass is already lower than other and we divide it by five since there is five which make it even lower --> high chance the leaves aren't all from this individual --> we may remove this individual later on

ggplot(data |>  filter(proj_2025 == "Traits"),
       aes(x = wet_mass_g, y = dry_mass_g)) +
  
  # scatter points
  geom_point(aes(colour = factor(plant_nr)),
             size = 2, alpha = 0.7) +
  
  # 1:1 reference line
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey50") +
  
  # fitted regression line
  geom_smooth(method = "lm", se = TRUE, colour = "blue") +
  
  facet_wrap(~ age_class) + # wet mass vs dry mass graphs facet by the age classes.
  
  # optional: label
  labs(
    x = "Wet mass per leaf (g)",
    y = "Dry mass per leaf (g)",
    title = "Dry mass vs wet mass per leaf",
    subtitle = "Dashed line = 1:1 reference, blue line = fitted linear trend",
    colour = "plant_nr"
  ) +
  
  theme_bw()

# for PIO 5 values are isolated 
# accordig to the dry mass graph it as a higher dry mass than others and a higher wet mass too 

```

# Calculate SLA and LDMC

**Specific Leaf Area (SLA) :**

SLA = Leaf Area (cm²) / Leaf dry mass (g)

High SLA (\> 300 000 mm²/g) means fast-growing plants ("Acquisitive strategy"). These plants have thin, flexible leaves that capture light efficiently.

Low SLA (\< 100 000 mm²/g) means slow-growing plants ("Conservative strategy"). These plants have thick, leathery leaves designed to last a long time and resist stress.

**Leaf Dry Matter Content (LDMC) :**

LDMC = Leaf Dry Mass (g)/ Leaf Fresh Mass (g)

Varies between 0 and 1

High LDMC (0.5 - 0.7) : Leaves are dense, high in lignin/cellulose, and low in water. They are very resistant to physical hazards and drought.

Low LDMC (0;05 - 0.5) : Leaves are "succulent" or watery.

```{r}
#| label: calculate-SLA-and-LDMC

data <- data |>  
  mutate(
    SLA = ifelse(
      !is.na(leaf_area) & !is.na(dry_mass), # excluding rows with NAs
      round(leaf_area / dry_mass, 3), # Leaf level metrics
      NA_real_
    )
  )

data <- data |> 
  mutate(
    LDMC = ifelse( # making new ldmc column
      !is.na(dry_mass) & !is.na(wet_mass), # excluding rows with NAs
      round(dry_mass / wet_mass, 5), # Leaf level metrics
      NA_real_
    )
  )
```

```{r}
#| label: check-SLA-outliers-(per-leaf)
#| warning: false

# making a scatterplot for dry mass mean for trait leaves, plant_nr on x-axis and weight in g on y-axis
ggplot(data |>  filter(proj_2025 == "Traits"),
       aes(x = factor(plant_nr), y = SLA, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "SLA",
    title = "Mean SLA per leaf by plant, age class and month"
  )

# If needed data modifications are documented in the adjusting-weird-values-in-the-data section.

## March
# BUI 6 (ADM0437) : 450, high --> 5 leaves at scanning, 5 leaves at dry mass, envelope checked

## May
# BUI 7 (ANH0353) : 110, low --> 4 leaves at scanning, 2 leaves at dry mass, envelope checked
# MAT 1 (ARA5864) : 567, high --> 4 leaves at scanning, 5 leaves at dry mass, envelope checked

## July
# NB 1 : high, checked
# PIO 1 (AUL1455) :high, --> 3 leaves at scanning, 3 leaves at dry mass, envelope checked.
```

```{r}
#| label: check-LDMC-outliers-(per-leaf)
#| warning: false

# making a scatterplot for dry mass mean for trait leaves, plant_nr on x-axis and weight in g on y-axis
ggplot(data |>  filter(proj_2025 == "Traits"),
       aes(x = factor(plant_nr), y = LDMC, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "Leaf Dry Matter Content (LDMC)",
    title = "Mean LDMC per leaf by plant, age class and month"
  )

# If needed data modifications are documented in the adjusting-weird-values-in-the-data section.

## March
# BUI 6 (ADM0437) : 0.265, low --> 5 leaves at dry mass, envelope checked
# MAT 6 (AEI1422) : 0.335, low --> 5 leaves at dry mass, envelope checked 

## May
# MAT 6 (ATQ5149) : 0.888, high, 4 leaves at wet mass and 4 leaves at dry mass, envelope checked
# BUI 2 (APL2797) : 1.720, too high. 4 leaves at wet mass (0.0043/leaf) and 2 leaves at dry mass (0.00074).Excluded sample.

```

# July Building 1 and 2 issue

The presence of a non-target branching leaf in BUI 1 and 2 during initial measurements led to artifactual trait discrepancies. Measurements were repeated 24 hours later to determine the appropriate dataset to retain.

```{r}
#| label: BUI-1-2-traits-check
#| warning: false

data |> 
  filter(
    month == "July",
    age_class == "Building",
    plant_nr %in% c("1", "2"),
    proj_2025 == "Traits"
  ) |> 
  pivot_longer(
    cols = c(LDMC, SLA, wet_mass, leaf_area, dry_mass),
    names_to = "trait",
    values_to = "value"
  ) |> 
  # Assign clean labels with units for each trait
  mutate(
    trait_label = case_when(
      trait == "dry_mass"  ~ "Dry Mass (g)",
      trait == "wet_mass"  ~ "Wet Mass (g)",
      trait == "leaf_area" ~ "Leaf Area (cm²)",
      trait == "LDMC"      ~ "LDMC (g/g)",
      trait == "SLA"       ~ "SLA (cm²/g)",
      TRUE ~ trait
    )
  ) |> 
  ggplot(
    aes(
      x = factor(plant_nr), 
      y = value, 
      color = case_when(
        grepl("harvested after 24h", remark) ~ "harvested after 24h",
        TRUE ~ "harvested like usual"
      )
    )
  ) +
  geom_point(size = 2, alpha = 0.8, position = position_jitter(width = 0.15)) +
  
  # Facet using the newly formatted label
  facet_wrap(~ trait_label, scales = "free_y") +
  
  scale_color_manual(
    name = "Harvest status",
    values = c(
      "harvested after 24h" = "firebrick",
      "harvested like usual" = "gray40"
    )
  ) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = NULL, # Remove y-axis label since units are now in the panel headers
    title = "Trait values per leaf by plant (July, Building)"
  )
```

```{r}
#| label: BUI-1-2-traits-check
#| warning: false


# 1. Prepare and transform the dataset
data_prepared <- data |> 
  filter(
    month == "July",
    age_class == "Building",
    proj_2025 == "Traits"
  ) |> 
  pivot_longer(
    cols = c(wet_mass, leaf_area, dry_mass, LDMC, SLA),
    names_to = "trait",
    values_to = "value"
  ) |> 
  mutate(
    trait_label = case_when(
      trait == "dry_mass"  ~ "Dry Mass (g)",
      trait == "wet_mass"  ~ "Wet Mass (g)",
      trait == "leaf_area" ~ "Leaf Area (cm²)",
      trait == "LDMC"      ~ "LDMC (g/g)",
      trait == "SLA"       ~ "SLA (cm²/g)",
      TRUE ~ trait
    ),
    harvest_status = case_when(
      grepl("harvested after 24h", remark) ~ "harvested after 24h",
      TRUE ~ "harvested like usual"
    )
  )

# 2. Loop over each unique trait_label and generate individual plots
unique(data_prepared$trait_label) |> 
  walk(function(t) {
    p <- data_prepared |> 
      filter(trait_label == t) |> 
      ggplot(aes(x = factor(plant_nr), y = value, color = harvest_status)) +
      
      geom_point(size = 2.5, alpha = 0.8, position = position_jitter(width = 0.15)) +
      
      scale_color_manual(
        name = "Harvest status",
        values = c(
          "harvested after 24h" = "firebrick",
          "harvested like usual" = "gray40"
        )
      ) +
      
      theme_bw() +
      labs(
        x = "Plant number",
        y = t,
        title = paste("Trait values per leaf by plant (July, Building) -", t)
      )
    
    print(p)
  })

# If needed data modifications are documented in the adjusting-weird-values-in-the-data section.

#Excluded samples: BUI 1 and 2 were measured twice: immediately and after 24 hours du to the presence of non-target branching leave in the first measurement. A  difference is observed between the two time points, with the 24-hour measurements being more consistent with the other buildings. Consequently, only the 24-hour measurements for BUI 1 and 2 were retained in the dataset.

data <- data |> 
  filter(
    !(
      month == "July" & 
      proj_2025 == "Traits" & 
      age_class == "Building" & 
      plant_nr %in% c("1", "2") & 
      (remark != "harvested after 24h" | is.na(remark))
    )
  )
```

# Check fluorescence outliers

```{r}
#| label: check-fluorescence-outliers
#| warning: false

# making a scatterplot for initial fluorescence
(data |> 
         filter(proj_2025 == "Traits") |> 
       ggplot(aes(x = factor(plant_nr), y = fv_fm_initial, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "fv_fm_initial",
    title = "Initial fluorescence by plant, age class and month"
  ))

(data |> 
         filter(proj_2025 == "TT Hot") |> 
       ggplot(aes(x = factor(plant_nr), y = fv_fm_final, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "fv_fm_final",
    title = "Final fluorescence TT Hot by plant, age class and month"
  ))

(data |> 
         filter(proj_2025 == "TT Cold") |> 
       ggplot(aes(x = factor(plant_nr), y = fv_fm_final, colour = factor(plant_nr))) +
  geom_point(size = 2, alpha = 0.8,
             position = position_jitter(width = 0.15)) +
  facet_grid(age_class ~ month) +
  theme_bw() +
  labs(
    x = "Plant number",
    y = "fv_fm_final",
    title = "Final fluorescence TT Cold by plant, age class and month"
  ))
# one value is registered as a January one --> modification is in data cleaning entire frame

#### Initial fluorescence

## March
# MAT 3 have a high fv/fm ; same for BUI 5 
# MAT and BUI low : brown leaves
# low points for NB 5, 7 and 8 --> maybe due to the season

## May 
# MAT 8 old : low --> brown leaves
# PIO 8 (AUS1188) : 0.38, low --> envelope checked

## July
# MAT 8 (BCP2114) : 0.01 envelope checked

#### final fluorescence TT Hot 
# High value for BUI 4, we don't have that in the initial fluo graph --> not done on the same leaves than initial fv/fm.
```

```{r}
#| label: check-fluorescence-outliers-TTHot
#| warning: false

# Filter initial dataset for TT Hot
data_filtered <- data |> 
  filter(proj_2025 == "TT Hot")

# Loop through each unique month to generate a distinct plot
unique(data_filtered$month) |> 
  walk(function(m) {
    p <- data_filtered |> 
      filter(month == m) |> 
      ggplot(aes(x = as.numeric(as.character(temp_exp)), 
                 y = fv_fm_final,
                 group = interaction(plant_nr, age_class),      
                 color = factor(plant_nr))) +
      
      geom_point(size = 2, alpha = 0.8,
                 position = position_jitter(width = 0.15)) +
      geom_line() +
      
      # Single-row panel grid for leaf_age (old vs young) and age_class
      facet_grid(age_class ~ leaf_age) + 
      
      theme_bw() +
      labs(
        x = "Temperature",
        y = "fv_fm_final",
        color = "plant_nr",
        title = paste("Fv/Fm response to hot temperature - Month:", m)
      )
    
    print(p)
  })

```

```{r}
#| label: check-fluorescence-outliers-TTCold
#| warning: false

# Filter initial dataset for TT Cold
data_filtered <- data |> 
  filter(proj_2025 == "TT Cold")

# Loop through each unique month to generate a distinct plot
unique(data_filtered$month) |> 
  walk(function(m) {
    p <- data_filtered |> 
      filter(month == m) |> 
      ggplot(aes(x = as.numeric(as.character(temp_exp)), 
                 y = fv_fm_final,
                 group = interaction(plant_nr, age_class),      
                 color = factor(plant_nr))) +
      
      geom_point(size = 2, alpha = 0.8,
                 position = position_jitter(width = 0.15)) +
      geom_line() +
      
      # Single-row panel grid for leaf_age (old vs young) and age_class
      facet_grid(age_class ~ leaf_age) + 
      
      theme_bw() +
      labs(
        x = "Temperature",
        y = "fv_fm_final",
        color = "plant_nr",
        title = paste("Fv/Fm response to cold temperature - Month:", m)
      )
    
    print(p)
  })

```

# Saving clean dataset

```{r}
#| label: clean-dataset


write.csv2(
  data, 
  file = "C:/Users/poula/Documents/M2_PAMEQ/Stage M2/Analyse statistique/thermotolerance/data/clean_data/Seasonal Calluna Thermal Tolerance and Plant Functional Traits.csv", 
  row.names = FALSE
)
```
